### 2025-03-13 paper/quarto_ecml/paper.tex

\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[misc]{ifsym}
\newcommand{\corr}{(\Letter)}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Fig.}
\else
  \newcommand\figurename{Fig.}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}

\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

To our surprise, the second challenge has not yet attracted major
research interest. Specifically, there has been no concerted effort
towards improving the ``explanatory capacity'' of models, i.e., the
degree to which learned representations correspond to explanations that
are \textbf{interpretable} and deemed \textbf{plausible} by humans (see
Definition~\ref{def-explainability}). Instead, the choice has generally
been to improve the ability of XAI tools to identify the subset of
explanations that are both plausible and valid for any given model,
independent of whether the learned representations are also compatible
with plausible explanations \cite{altmeyer2024faithful}. Fortunately,
recent findings indicate that improved explanatory capacity can arise as
a consequence of regularization techniques aimed at other training
objectives such as robustness, generalization, and generative capacity
\cite{schut2021generating,augustin2020adversarial,altmeyer2024faithful}.
As further discussed in Section~\ref{sec-lit}, our work consolidates
these findings within a single objective.

@@

\begin{definition}[Model
Explainability]\protect\hypertarget{def-explainability}{}\label{def-explainability}

Let \(\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}\) denote a
supervised classification model that maps from the \(D\)-dimensional
input space \(\mathcal{X}\) to representations
\(\phi(\mathbf{x};\theta)\) and finally to the \(K\)-dimensional output
space \(\mathcal{Y}\). Assume that for any given input-output pair
\(\{\mathbf{x},\mathbf{y}\}_i\) there exists a counterfactual
\(\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})\)
where \(\arg\max_y{\mathbf{y}^{+}}=y^+\) and \(y^+\) denotes the index
of the target class.

We say that \(\mathbf{M}_\theta\) is \textbf{explainable} to the extent
that faithfully generated counterfactuals are plausible (i.e.,
consistent with the data) and actionable. Formally, we define these
properties as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  (Plausibility)
  \(\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1\)
  where \(A\) is some small region around \(\mathbf{x}^{\prime}\).
\item
  (Actionability) Permutations \(\Delta\) are subject to some
  actionability constraints.
\end{enumerate}

We consider counterfactuals as faithful to the extent that they are
consistent with what the model has learned about the input data. Let
\(p_\theta(\mathbf{x}|\mathbf{y}^{+})\) denote the conditional posterior
over inputs, then formally:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  (Faithfulness)
  \(\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1\)
  where \(A\) is defined as above.
\end{enumerate}

\end{definition}

The characterization of faithfulness and plausibility in
Definition~\ref{def-explainability} is the same as in
\cite{altmeyer2024faithful}, with adapted notation. Actionability
constraints in Definition~\ref{def-explainability} vary and depend on
the context in which \(\mathbf{M}_\theta\) is deployed. In this work, we
focus on domain and mutability constraints for individual features
\(x_d\) for \(d=1,...,D\). We limit ourselves to classification tasks
for reasons discussed in Section~\ref{sec-discussion}.

@@

Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
explanation generated through gradient descent over \(T\) iterations as
initially proposed by \cite{wachter2017counterfactual}. For our
purposes, we let \(T\) vary and consider the counterfactual search to be
converged as soon as the predicted probability for the target class has
reached a pre-determined threshold \(\tau\):
\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
where \(\mathcal{S}\) is the softmax function.\footnote{For detailed
  background information on gradient-based counterfactual search and
  convergence see \textbf{?@sec-app-ce}.}

To train models with high explainability as defined in
Definition~\ref{def-explainability}, we propose to leverage
counterfactuals in the following objective:

@@

A key empirical finding in \cite{altmeyer2024faithful} was that JEMs
tend to do well with respect to the plausibility objective in
Definition~\ref{def-explainability}. This follows directly if we
consider samples drawn from \(p_\theta(\mathbf{x})\) as counterfactuals
because the JEM objective effectively minimizes the divergence between
the conditional posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\). To
generate samples, \cite{grathwohl2020your} rely on Stochastic Gradient
Langevin Dynamics (SGLD) using an uninformative prior for initialization
but we depart from their methodology. Instead of SGLD, we propose to use
counterfactual explainers to generate counterfactuals of observed
training samples. Specifically, we have:

@@

Intuitively, the gradient of Equation~\ref{eq-div} decreases the energy
of observed training samples (positive samples) while increasing the
energy of counterfactuals (negative samples) \cite{du2019implicit}. As
the counterfactuals get more plausible
(Definition~\ref{def-explainability}) during training, these opposing
effects gradually balance each other out \cite{lippe2024uvadlc}.

@@

\begin{proposition}[Protecting Immutable
Features]\protect\hypertarget{prp-mtblty}{}\label{prp-mtblty}

Let
\(f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})\)
denote a linear classifier with softmax activation \(\mathcal{S}\)
(i.e., \emph{multinomial logistic regression}) where
\(y\in\{1,...,K\}=\mathcal{K}\) and \(\mathbf{x} \in \mathbb{R}^D\). If
we assume multivariate Gaussian class densities with common diagonal
covariance matrix \(\Sigma_k=\Sigma\) for all \(k \in \mathcal{K}\),
then protecting an immutable feature from the contrastive divergence
penalty (Equation~\ref{eq-div}) will result in lower classifier
sensitivity to that feature relative to the remaining features, provided
that at least one of those is discriminative and mutable.

\end{proposition}

It is worth highlighting that Proposition~\ref{prp-mtblty} assumes
independence of features. This raises a valid concern about the effect
of protecting immutable features in the presence of proxies that remain
unprotected. We address this in Section~\ref{sec-discussion}.

\subsection{Example: influence of counterfactual training on
explainability}\label{example-influence-of-counterfactual-training-on-explainability}

To better convey the intuition underlying our proposed method, we
illustrate different model outcomes in Example~\ref{exm-grad}.

\begin{example}[Prediction of Consumer Credit
Default]\protect\hypertarget{exm-grad}{}\label{exm-grad}

@@

In panel (a), we have trained our model conventionally, and we do not
impose mutability constraints at test time. The generated
counterfactuals are all valid, but not plausible: they do not comply
with the distribution of the factual samples in the target class to the
point where they are clearly distinguishable from the ground-truth data.
In panel (b), we have trained our model with counterfactual training,
once again without any mutability constraints. We observe that the
counterfactuals are highly plausible, meeting the first objective of
Definition~\ref{def-explainability}.

In panel (c), we have used conventional training again, this time
imposing the mutability constraint on \emph{age} at test time.
Counterfactuals are valid but involve some substantial reductions in
\emph{debt} for some individuals (very young applicants). By comparison,
counterfactual paths are shorter on average in panel (d), where we have
used counterfactual training and protected the immutable feature as
described in Section~\ref{sec-constraints}. We observe that due to the
classifier's lower sensitivity to \emph{age}, recourse recommendations
with respect to \emph{debt} are much more homogenous and do not
disproportionately punish younger individuals. The counterfactuals are
also plausible with respect to the mutable feature. Thus, we consider
the model in panel (d) as the most explainable according to
Definition~\ref{def-explainability}.

\end{example}

@@

\caption{\label{fig-poc}Illustration for Example~\ref{exm-grad} of how
CT improves model explainability.}

@@

\begin{corollary}[Plausibility]\protect\hypertarget{cor-plaus}{}\label{cor-plaus}

To what extent does our proposed counterfactual training objective
(Equation~\ref{eq-obj}) induce models to learn plausible explanations?

\end{corollary}

\begin{corollary}[Actionability]\protect\hypertarget{cor-action}{}\label{cor-action}

To what extent does our proposed counterfactual training objective
(Equation~\ref{eq-obj}) yield more favorable algorithmic recourse
outcomes in the presence of actionability constraints?

\end{corollary}

Beyond this, we are also interested in understanding how robust our
answers to RQ~\ref{cor-plaus} and RQ~\ref{cor-action} are:

\begin{corollary}[Hyperparameters]\protect\hypertarget{cor-hyper}{}\label{cor-hyper}

What are the effects of hyperparameter selection with respect to
Equation~\ref{eq-obj}?

\end{corollary}

@@

Our key outcome of interest is how well do models perform with respect
to explainability (Definition~\ref{def-explainability}). To this end, we
focus primarily on the plausibility and cost of faithfully generated
counterfactuals at test time. To measure the cost of counterfactuals, we
follow the standard convention of using distances (\(\ell_1\)-norm)
between factuals and counterfactuals as a proxy. For plausibility, we
assess how similar counterfactuals are to observed samples in the target
domain. We rely on the distance-based metric used by
\cite{altmeyer2024faithful},

\begin{equation}\phantomsection\label{eq-impl-dist}{
\text{implaus}_{\text{dist}}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-impl-div}{
\text{implaus}_{\text{div}}(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
}\end{equation}

@@

\subsubsection{Plausibility}\label{plausibility-1}

\subsubsection{Actionability}\label{actionability-1}

@@

\textbf{\emph{CT increases the training time of models.}} Counterfactual
training promotes explainability through CEs and robustness through AEs
at the cost of longer training times compared to conventional training
regimes. While higher numbers of iterations and counterfactuals per
iteration positively impact the quality of found solutions, they also
increase the required amount of computations. We find that relatively
small grids with 270 settings can take almost four hours for more
demanding datasets on a high-performance computing cluster with 34 2GB
CPUs (see details in \textbf{?@sec-hardware}). However, there are three
factors that attenuate the impact of this limitation. First, CT provides
CEs for the training samples for free, which may be beneficial in many
ADM systems. Second, we find that CT can retain its value when used as a
``fine-tuning'' training regime for conventionally-trained models.
Third, in principle, CT yields itself to parallel execution, which we
have leveraged for our own experiments.

@@

\newpage{}

>>>>

\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[misc]{ifsym}
\newcommand{\corr}{(\Letter)}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Fig.}
\else
  \newcommand\figurename{Fig.}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

To our surprise, the second challenge has not yet attracted major
research interest. Specifically, there has been no concerted effort
towards improving the ``explanatory capacity'' of models, i.e., the
degree to which learned representations correspond to explanations that
are \textbf{interpretable} and deemed \textbf{plausible} by humans (see
Def. \ref{def-explainability}). Instead, the choice has generally been
to improve the ability of XAI tools to identify the subset of
explanations that are both plausible and valid for any given model,
independent of whether the learned representations are also compatible
with plausible explanations \cite{altmeyer2024faithful}. Fortunately,
recent findings indicate that improved explanatory capacity can arise as
a consequence of regularization techniques aimed at other training
objectives such as robustness, generalization, and generative capacity
\cite{schut2021generating,augustin2020adversarial,altmeyer2024faithful}.
As further discussed in Section~\ref{sec-lit}, our work consolidates
these findings within a single objective.

@@

\begin{definition}[Model Explainability]
\label{def-explainability}
Let $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denote a supervised classification model that maps from the $D$-dimensional input space $\mathcal{X}$ to representations $\phi(\mathbf{x};\theta)$ and finally to the $K$-dimensional output space $\mathcal{Y}$. Assume that for any given input-output pair $\{\mathbf{x},\mathbf{y}\}_i$ there exists a counterfactual $\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})$ where $\arg\max_y{\mathbf{y}^{+}}=y^+$ and $y^+$ denotes the index of the target class.

We say that $\mathbf{M}_\theta$ is \textbf{explainable} to the extent that faithfully generated counterfactuals are plausible and actionable. Formally, we define these properties as follows,

\begin{enumerate}
    \item (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is some small region around $\mathbf{x}^{\prime}$.
    \item (Actionability) Permutations $\Delta$ are subject to some actionability constraints.
    \item (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is defined as above.
\end{enumerate}

where $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior over inputs. 
\end{definition}

The characterization of faithfulness and plausibility in Def.
\ref{def-explainability} is the same as in \cite{altmeyer2024faithful},
with adapted notation. Intuitively, plausible counterfactuals are
consistent with the data and faithful counterfactuals are consistent
with what the model has learned about input data. Actionability
constraints in Def. \ref{def-explainability} vary and depend on the
context in which \(\mathbf{M}_\theta\) is deployed. In this work, we
focus on domain and mutability constraints for individual features
\(x_d\) for \(d=1,...,D\). We limit ourselves to classification tasks
for reasons discussed in Section~\ref{sec-discussion}.

@@

Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
explanation generated through gradient descent over \(T\) iterations as
initially proposed by \cite{wachter2017counterfactual}. For our
purposes, we let \(T\) vary and consider the counterfactual search as
converged as soon as the predicted probability for the target class has
reached a pre-determined threshold, \(\tau\):
\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
where \(\mathcal{S}\) is the softmax function.\footnote{For detailed
  background information on gradient-based counterfactual search and
  convergence see supplementary appendix.}

To train models with high explainability as defined in Def.
\ref{def-explainability}, we propose to leverage counterfactuals in the
following objective:

@@

A key empirical finding in \cite{altmeyer2024faithful} was that JEMs
tend to do well with respect to the plausibility objective in Def.
\ref{def-explainability}. This follows directly if we consider samples
drawn from \(p_\theta(\mathbf{x})\) as counterfactuals because the JEM
objective effectively minimizes the divergence between the conditional
posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\). To generate samples,
\cite{grathwohl2020your} rely on Stochastic Gradient Langevin Dynamics
(SGLD) using an uninformative prior for initialization but we depart
from their methodology. Instead of SGLD, we propose to use
counterfactual explainers to generate counterfactuals of observed
training samples. Specifically, we have:

@@

Intuitively, the gradient of Equation~\ref{eq-div} decreases the energy
of observed training samples (positive samples) while increasing the
energy of counterfactuals (negative samples) \cite{du2019implicit}. As
the counterfactuals get more plausible (Def. \ref{def-explainability})
during training, these opposing effects gradually balance each other out
\cite{lippe2024uvadlc}.

@@

\begin{proposition}[Protecting Immutable Features]
\label{prp-mtblty}
Let $f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})$ denote a linear classifier with softmax activation $\mathcal{S}$ where $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$. If we assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty will result in lower classifier sensitivity to that feature relative to the remaining features, provided that at least one of those is discriminative and mutable.
\end{proposition}

It is worth highlighting that Prp.\textasciitilde{}\ref{prp-mtblty}
assumes independence of features. This raises a valid concern about the
effect of protecting immutable features in the presence of proxies that
remain unprotected. We address this in Section~\ref{sec-discussion}.

\subsection{Example (Prediction of Consumer Credit
Default)}\label{example-prediction-of-consumer-credit-default}

@@

In panel (a), we have trained our model conventionally, and we do not
impose mutability constraints at test time. The generated
counterfactuals are all valid, but not plausible: they do not comply
with the distribution of the factual samples in the target class to the
point where they are clearly distinguishable from the ground-truth data.
In panel (b), we have trained our model with counterfactual training,
once again without any mutability constraints. We observe that the
counterfactuals are highly plausible, meeting the first objective of
Def. \ref{def-explainability}.

In panel (c), we have used conventional training again, this time
imposing the mutability constraint on \emph{age} at test time.
Counterfactuals are valid but involve some substantial reductions in
\emph{debt} for some individuals (very young applicants). By comparison,
counterfactual paths are shorter on average in panel (d), where we have
used counterfactual training and protected the immutable feature as
described in Section~\ref{sec-constraints}. We observe that due to the
classifier's lower sensitivity to \emph{age}, recourse recommendations
with respect to \emph{debt} are much more homogenous and do not
disproportionately punish younger individuals. The counterfactuals are
also plausible with respect to the mutable feature. Thus, we consider
the model in panel (d) as the most explainable according to Def.
\ref{def-explainability}.

@@

\caption{\label{fig-poc}Illustration of how CT improves model
explainability.}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  To what extent does our proposed counterfactual training objective
  (Equation~\ref{eq-obj}) induce models to learn plausible explanations?
\item
  To what extent does our proposed counterfactual training objective
  (Equation~\ref{eq-obj}) yield more favorable algorithmic recourse
  outcomes in the presence of actionability constraints?
\item
  What are the effects of hyperparameter selection with respect to
  Equation~\ref{eq-obj}?
\end{enumerate}

@@

Our key outcome of interest is how well do models perform with respect
to explainability (Def. \ref{def-explainability}). To this end, we focus
primarily on the plausibility and cost of faithfully generated
counterfactuals at test time. To measure the cost of counterfactuals, we
follow the standard convention of using distances (\(\ell_1\)-norm)
between factuals and counterfactuals as a proxy. For plausibility, we
assess how similar counterfactuals are to observed samples in the target
domain. We rely on the distance-based metric used by
\cite{altmeyer2024faithful},

\begin{equation}\phantomsection\label{eq-impl-dist}{
\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-impl-div}{
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
}\end{equation}

@@

\subsubsection{Plausibility}\label{plausibility}

Table~\ref{tbl-main} presents our main empirical findings. The top five
rows show the percentage reduction in implausibility according to
Equation~\ref{eq-impl-dist} for varying degrees of the energy penalty
used for \emph{ECCo} at test time. The following row shows the reduction
in implausibility as measured by Equation~\ref{eq-impl-div} and
aggregated across all test specifications of \emph{ECCo}. The final two
rows show the test accuracies for the model trained with CT and
conventionally trained models (``vanilla'').

We observe that for all datasets except \emph{OL} and across all test
settings, the average distance of counterfactuals from observed samples
in the target class is reduced, indicating improved plausibility. The
magnitude of improvements varies by dataset: for the simple synthetic
datasets, distance reductions range from around 20-40\% (\emph{LS},
\emph{Moon}) to almost 60\% (\emph{Circ}). For the real-world tabular
datasets, improvements are generally smaller but still substantial in
many cases with around 10-15\% for \emph{CH}, 11-28\% for \emph{GMSC},
7-8\% for \emph{Cred} and around 3\% for \emph{Adult}. For our only
vision dataset (\emph{MNIST}), distances are reduced by up to 9\%. The
results for our proposed divergence metric are qualitatively similar,
but generally even more pronounced: for the \emph{Circ} dataset,
implausibility is reduced by almost 94\% to virtually zero as we
verified by looking at the absolute outcome. Improvements for other
datasets range from 28\% (\emph{Moon}) to 78\% (\emph{GMSC}). For
\emph{OL} the reduction is negative, consistent with the distance-based
metric. The only dataset, for which our proposed metric disagrees with
the distance-based metric is \emph{MNIST}.

These broad and substantial improvements in plausibility generally do
not come at the cost of decreased predictive performance: test accuracy
for CT is virtually identical to the baseline for \emph{Adult},
\emph{Circ}, \emph{LS}, \emph{Moon} and \emph{OL}, and even slightly
improved for \emph{Cred}. Exceptions to this general pattern are
\emph{MNIST}, \emph{CH} and \emph{GMSC}, for which we observe reduction
in test accuracy of 2, 5 and 15 percentage points, respectively. We note
in this context, that we have not optimized our models for predictive
performance at all and worked with very small networks. In summary, we
find that CT can substantially improve the quality of explanations
learned by models without generally sacrificing predictive accuracy.

\begin{table}

\caption{\label{tbl-main}Key plausibility and predictive performance
metrics for all datasets. The top five rows show the percentage
reduction in implausibility according to Equation~\ref{eq-impl-dist} for
varying degrees of the energy penalty used for \emph{ECCo} at test time.
The following row shows the reduction in implausibility as measured by
Equation~\ref{eq-impl-div} and aggregated across all test specifications
of \emph{ECCo}. The final two rows show the test accuracies for the
model trained with CT and conventionally trained models (``vanilla'').}

\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}} cccccc }
\input{tables/main.tex}
\end{tabular*}

\end{table}%

\subsubsection{Actionability}\label{actionability}

@@

\textbf{\emph{CT increases the training time of models.}} Counterfactual
training promotes explainability through CEs and robustness through AEs
at the cost of longer training times compared to conventional training
regimes. While higher numbers of iterations and counterfactuals per
iteration positively impact the quality of found solutions, they also
increase the required amount of computations. We find that relatively
small grids with 270 settings can take almost four hours for more
demanding datasets on a high-performance computing cluster with 34 2GB
CPUs\footnote{See supplementary appendix for computational details.}.
However, there are three factors that attenuate the impact of this
limitation. First, CT provides counterfactual explanations for the
training samples essentially for free, which may be beneficial in many
ADM systems. Second, we find that CT can retain its value when used as a
``fine-tuning'' training regime for conventionally-trained models.
Third, in principle, CT yields itself to parallel execution, which we
have leveraged for our own experiments.

@@

### 2025-03-13 paper/sections/appendix.qmd

\setcounter{section}{0}
\renewcommand{\thesection}{\Alph{section}}

\setcounter{table}{0}
\renewcommand{\thetable}{A\arabic{table}}

\setcounter{figure}{0}
\renewcommand{\thefigure}{A\arabic{figure}}

@@

# Technical Details of Our Approach {.appendix}

{{< include /paper/sections/other/counterfactuals.qmd >}}

{{< include /paper/sections/other/constraints.qmd >}}

{{< include /paper/sections/other/training.qmd >}}

# Detailed Results {.appendix}

## Qualitative Findings for Image Data

::: {.callout-note}

@fig-mnist shows much more plausible (faithful) counterfactuals for a model with CT than the model with conventional training (@fig-mnist-vanilla). In fact, this is not even using *ECCo+* and still showing better results than the best results we achieved in our AAAI paper for JEM ensembles.

:::

![Counterfactual images for *MLP* with counterfactual training. The underlying generator, *ECCo*, aims to generate counterfactuals that are faithful to the model [@altmeyer2024faithful].](/paper/figures/mnist_mlp.png){#fig-mnist}

![Counterfactual images for *MLP* with conventional training. The underlying generator, *ECCo*, aims to generate counterfactuals that are faithful to the model [@altmeyer2024faithful].](/paper/figures/mnist_mlp_vanilla.png){#fig-mnist-vanilla}

@@

<!-- {{< include /paper/sections/other/initial_grids.qmd >}} -->

>>>>

<!-- \setcounter{section}{0} -->
\renewcommand{\thesection}{\Alph{section}}

<!-- \setcounter{table}{0} -->
\renewcommand{\thetable}{A\arabic{table}}

<!-- \setcounter{figure}{0} -->
\renewcommand{\thefigure}{A\arabic{figure}}

@@

{{< include /paper/sections/other/technical.qmd >}}

{{< include /paper/sections/other/main_results.qmd >}}

@@

{{< include /paper/sections/other/tuning.qmd >}}

### 2025-03-13 paper/sections/discussion.qmd

***CT increases the training time of models.*** Counterfactual training promotes explainability through CEs and robustness through AEs at the cost of longer training times compared to conventional training regimes. While higher numbers of iterations and counterfactuals per iteration positively impact the quality of found solutions, they also increase the required amount of computations. We find that relatively small grids with 270 settings can take almost four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs (see details in @sec-hardware). However, there are three factors that attenuate the impact of this limitation. First, CT provides CEs for the training samples for free, which may be beneficial in many ADM systems. Second, we find that CT can retain its value when used as a "fine-tuning" training regime for conventionally-trained models. Third, in principle, CT yields itself to parallel execution, which we have leveraged for our own experiments.

>>>>

***CT increases the training time of models.*** Counterfactual training promotes explainability through CEs and robustness through AEs at the cost of longer training times compared to conventional training regimes. While higher numbers of iterations and counterfactuals per iteration positively impact the quality of found solutions, they also increase the required amount of computations. We find that relatively small grids with 270 settings can take almost four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs^[See supplementary appendix for computational details.]. However, there are three factors that attenuate the impact of this limitation. First, CT provides counterfactual explanations for the training samples essentially for free, which may be beneficial in many ADM systems. Second, we find that CT can retain its value when used as a "fine-tuning" training regime for conventionally-trained models. Third, in principle, CT yields itself to parallel execution, which we have leveraged for our own experiments.

### 2025-03-13 paper/sections/experiments.qmd

::: {#cor-plaus}

## Plausibility

Does our proposed counterfactual training objective (@eq-obj) induce models to learn plausible explanations?

::: {#cor-action}

## Actionability

Does our proposed counterfactual training objective (@eq-obj) yield more favorable algorithmic recourse outcomes in the presence of actionability constraints?

Beyond this, we are also interested in understanding how robust our answers to @cor-plaus and @cor-action are:

::: {#cor-hyper}

## Hyperparameters

What are the effects of different hyperparameter choices with respect to @eq-obj?

@@

Our key outcome of interest is how well models perform with respect to explainability (@def-explainability): to this end, we focus primarily on the plausibility and cost of faithfully generated counterfactuals at test time. To measure the cost of counterfactuals, we follow the standard convention of using distances ($\ell_1$-norm) between factuals and counterfactuals as a proxy. For plausibility, we assess how similar counterfactuals are to observed samples in the target domain. We rely on the distance-based metric used by @altmeyer2024faithful,

>>>>

1. To what extent does our proposed counterfactual training objective (@eq-obj) induce models to learn plausible explanations?
2. To what extent does our proposed counterfactual training objective (@eq-obj) yield more favorable algorithmic recourse outcomes in the presence of actionability constraints?
3. What are the effects of hyperparameter selection with respect to @eq-obj?

@@

Our key outcome of interest is how well do models perform with respect to explainability (Def. \ref{def-explainability}). To this end, we focus primarily on the plausibility and cost of faithfully generated counterfactuals at test time. To measure the cost of counterfactuals, we follow the standard convention of using distances ($\ell_1$-norm) between factuals and counterfactuals as a proxy. For plausibility, we assess how similar counterfactuals are to observed samples in the target domain. We rely on the distance-based metric used by @altmeyer2024faithful,

### 2025-03-13 paper/sections/intro.qmd

To our surprise, the second challenge has not yet attracted major research interest. Specifically, there has been no concerted effort towards improving the "explanatory capacity" of models, i.e., the degree to which learned representations correspond to explanations that are **interpretable** and deemed **plausible** by humans (see @def-explainability). Instead, the choice has generally been to improve the ability of XAI tools to identify the subset of explanations that are both plausible and valid for any given model, independent of whether the learned representations are also compatible with plausible explanations [@altmeyer2024faithful]. Fortunately, recent findings indicate that improved explanatory capacity can arise as a consequence of regularization techniques aimed at other training objectives such as robustness, generalization, and generative capacity [@schut2021generating; @augustin2020adversarial; @altmeyer2024faithful]. As further discussed in @sec-lit, our work consolidates these findings within a single objective.

>>>>

To our surprise, the second challenge has not yet attracted major research interest. Specifically, there has been no concerted effort towards improving the "explanatory capacity" of models, i.e., the degree to which learned representations correspond to explanations that are **interpretable** and deemed **plausible** by humans (see Def. \ref{def-explainability}). Instead, the choice has generally been to improve the ability of XAI tools to identify the subset of explanations that are both plausible and valid for any given model, independent of whether the learned representations are also compatible with plausible explanations [@altmeyer2024faithful]. Fortunately, recent findings indicate that improved explanatory capacity can arise as a consequence of regularization techniques aimed at other training objectives such as robustness, generalization, and generative capacity [@schut2021generating; @augustin2020adversarial; @altmeyer2024faithful]. As further discussed in @sec-lit, our work consolidates these findings within a single objective.
